Pervasive nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2025-04-05 manuscript/manuscript_file.Rmd
Minor grammar corrections following Susan Cormier's review

**Lakes and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere but recent evidence of N~2~O uptake across a broad range of lakes have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was pervasive through the CONUS, with an estimated `r national_stats$post_m_prop_sat * 100`% (95% credible interval: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) of lakes functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate concentration, waterbody surface area, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS lakes is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Pervasive N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS lakes are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

Despite pervasive N~2~O undersaturation across the POI, estimated means for N~2~O saturation ratio (ratio of measured to equilibrium concentration) were indistinguishable from a value of one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction of pervasive N~2~O undersaturation and average dissolved N~2~O concentrations close to equilibrium can be attributed to the prominent right skew in the predicted distribution of dissolved N~2~O concentrations across the POI **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, the predicted distribution of N~2~O saturation ratios had a median value less than one, indicating that most lakes in these populations were undersaturated . However, the long right tails on these distributions, comprised of relatively rare lakes with very high levels of N~2~O supersaturation, were sufficient to drive the population means to values close to, or even greater than one. Thus, most lakes in the CONUS were functioning as N~2~O sinks during the summer months, but a few functioned as strong N~2~O sources, particularly at high NO~3~ and high water temperature **(Fig. 3)**.

>>>>

**Lakes and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere but recent evidence of N~2~O uptake across a broad range of lakes have called the accuracy of emission estimates into question. Here we use a new national-scale dataset of dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was pervasive throughout the CONUS, with an estimated `r national_stats$post_m_prop_sat * 100`% (95% credible interval: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) of lakes functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate concentration, waterbody surface area, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS lakes is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Pervasive N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS lakes are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

Despite pervasive N~2~O undersaturation across the POI, estimated means for N~2~O saturation ratio (ratio of measured to equilibrium concentration) were indistinguishable from a value of one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction of pervasive N~2~O undersaturation and average dissolved N~2~O concentrations close to equilibrium can be attributed to the prominent right skew in the predicted distribution of dissolved N~2~O concentrations across the POI **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, the predicted distribution of N~2~O saturation ratios had a median value less than one, indicating that most lakes in these populations were undersaturated. However, the long right tails on these distributions, comprised of relatively rare lakes with very high levels of N~2~O supersaturation, were sufficient to drive the population means to values close to, or even greater than one. Thus, most lakes in the CONUS were functioning as N~2~O sinks during the summer months, but a few functioned as strong N~2~O sources, particularly at high NO~3~ and high water temperature **(Fig. 3)**.
